Bees need larger brains to thrive in urban environmnets

### 2023-02-05 Manuscript/Ms.Rmd
Preparing the MS for biology letters, formatting on length, number of figures and adding pertinent refs

---
title: "Urban bees have larger brains"
output:
  pdf_document:
    fig_caption: yes

@@

header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float} 
- \floatplacement{figure}{H} 
---

@@

[Not a great fan of the title, maybe something like this?](J)
[Bees with larger brains relative to their body are more common in urban environments](J)

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role in ecosystems for fruit and seed production. However, not all pollinators are negatively affected by habitat conversion, but some find in anthropogenic landscapes appropriate resources to survive and reproduce. Why some animals can thrive in anthropogenic environments that most species cannot tolerate remains insufficiently understood, but it might in part reflect enhanced behavioural plasticity to exploit new resource opportunities. While this hypothesis has received ample support in studies of vertebrates, it is less obvious whether the small brain of insects can offer enough behavioural plasticity to cope with the many challenges of anthropogenic habitats. To address this issue, we measured brains for XXX individuals from XX species of bees, the most important group of pollinators, and compared them with detailed information on habitat preferences. Our analyses revealed that bees that prefer forested or agricultural habitats had smaller brains relative to their body size, but those who prefer urban habitats tended to have larger relative brain size. In addition, urban bees also tend to be larger, and hence, show larger absolute brain sizes. These results suggest that a large brain can help maintaining bee urban populations.

Keywords: relative brain size, habitat preference, Apoidea, Urbanization.

Here, we characterize habitat preference for XXX bee species [@collado2019] and relate this to their brain size, expecting to find bigger brain sizes for urban dwellers because a wider array of cognitive abilities are needed to inhabit these habitats. In vertebrates, relative brain size is a good proxy of problem-solving and learning (REF!!), which can help individuals to cope with the many challenges on novel environments. Despite there was no unequivocal evidence that insect brain size is related to cognitive performance, learning abilities across bee species are reflected in differences in their brain size [@collado2021].  Hence, our study explores the long standing idea that species with higher encephalisation (i.e., large brain mass relative to their body) are more likely to thrive in urbanised areas applies also to insects.

## Overview

[yo no pondria refs en el parrafo de overview](J)

To evaluate our main hypothesis, we have used a dataset of bee species captured in North America and Europe with measurements of brain mass and body size [@sayol2020]. For all these species, we downloaded the occurrence records from the Global Biodiversity Information Facility (GBIF). After filtering the data to avoid possible biases, we extracted the land use classification of the different georeferenced records and summarised them into 3 different categories: (1) natural, (2) agricultural and (3) urban. To evaluate if the different bee species occur more in some habitats than others, we calculated their habitat preference by comparing the observed occurrences with the ones of null models assuming random distributions across habitats [@collado2019]. Finally, for the species that showed low and high statistical preference for the different habitats, we investigated how the association between habitat preference and relative brain size changed by habitat type.

All analyses were undertaken in R version 4.0.5 [@r2021] and the manuscript was written in rmarkdown version 2.17 [@markdown2022]. All data processing and graphics were done with the set of packages from the tidyverse version 1.3.0 [@tidyverse19].

@@

@@

Our dataset contains measurements of brain and body size for bee specimens captured by hand netting in different areas of the East Coast of United States and Europe (i.e., Spain and Netherlands). In total, the dataset includes `r sum(d$individuals)` female individuals from `r nlevels(factor(unique(d$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` different families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) females are involved in a larger number of tasks than males and are likely to experience greater selective pressures from the environment; and (ii), there are evident behavioral and morphological differences between sexes that are associated with the size and structure of the brain [@streinzer2013; @roselino2015]. In addition, because brain size can change with age and experience of the different individuals within a species [@durst1994; @withers2008], we limited our sampling to specimens foraging on flowers. Note that 93 of the species used in our study were previously published with their respective trait measurements in @sayol2020.

## (b) Brain and body size measurements

Brain size was measured as the weight of the fixed brain for each specimen. First, all collected individuals were kept in cold and sent to the laboratory where they were identified by expert taxonomists and anesthetized in cold [@crook2013]. Then, heads were removed with a scalpel and placed in a fixative solution (4% paraformaldehyde in 0.1 M phosphate-buffered saline -PBS- pH = 7.4) to avoid degradation. Subsequently, brains were extracted from the head capsule and placed on a petri dish after removing the retina from the optic lobes and cleaning the entire brain of all tracheae and fat bodies. By doing this, we ensured that our brain measurements mainly accounted for neural tissue. We then placed the brains on a small piece of Parafilm® and dried the exceeding fixative solution with Kimwipes® tissues. Then, within 4 seconds of the liquid removal, we weighted the brains with the help of a microbalance Sartorius Cubis®. Finally, we calculated the average weight for the species with more than one specimen. Because weighing bee brains is a highly error-prone task, we excluded values that were under or above 1.5 times the interquartile range for each species.

We considered as a proxy of body size the average intertegular span (ITS) per species. ITS represents the distance between the base of the wings (tegulae) on the bee thorax and has been shown to accurately predict body size as it is highly correlated with dry body weight [@cane1987; @kendall2019]. All ITS measurements were conducted with a stereomicroscope (magnification 16x or 80x) with a calibrated ocular micrometre (resolution down to 0.02 mm).

## (c) Occurrence data

@@

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. We selected the boundaries from the maps packages *spData* version 2.2.0 [@spdata] and *giscoR* vesion 0.3.2 [@giscoR], and extracted the occurrences with the help of the function *st_intersection* from the package *sf* version 1.0-7 [@pebesma2018]. For North America, we selected states located on the East Coast of United States (Figure S1A) that covered an approximate area of `r usakm2` km$^{2}$. These states were: Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. Similarly, we selected countries located on the north and center of Europe (Figure S1B) that represented a total area of `r eukm2` km$^{2}$. These countries were: England, Belgium, Netherlands, Germany and Denmark. All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.5s5kuf where just the countries of interest were retrieved. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a distribution larger than the area selected; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large differences with the extracted land cover classifications, which is from 2006. After pruning the data, our final dataset comprised `r spp_number` species with  `r usa_occurrences` occurrence records for United States (`r usa_spp` species; median occurrences = `r usa_median`) and `r eu_occurrences` records for Europe (`r eu_spp` species; median occurrences = `r eu_median`) wit a total of `r spp_shared` shared species between both regions.

## (d) Land cover classification

@@

## (e) Construction of the phylogenetic tree

@@

## (f) Analysis

We estimated brain size relative to body size modeling the log-transformed association between brain weight and intertegular span. For this, we specified a simple Bayesian regression with the function *brm* from the package *brms* version 2.18 [@burkner2017] where brain weight was the response variable and intertegular span the predictor. To correct for the non-independent evolutionary histories among species, we included the phylogenetic covariance matrix as random factor. Note that this matrix contained all species of our dataset (N = `r spp_number_nofilter`) because we expect better predictive power with a larger sample size than the one with the filtered dataset (N = `r spp_number`). Lastly, we extracted the differences between the observed and predicted values (i.e., residuals) that represent the relative brain size. High values indicate larger brains than expected for their body size and low values denote smaller brains than expected.

@@

@@

Bees with a preference for inhabiting urban habitats tended to have bigger brains related to their body sizes, and those with forest or agricultural  preferences had relatively smaller brains. This pattern is highly suggestive, because investing in brain tissue is expensive (REF), and hence it only payoff in unpredictable habitats where a flexible behavior is needed. When the environment is stable, small relative brains are cost-effective to exploit predictable resources, but in highly unpredictable environments, a large brain can confer the needed flexibility to exploit unpredictable resources. Indeed, specialist bees also tend to have larger brains [@sayol2020], as they need to locate and memorize patchy resources. This same complex learning and navigation skills may be also needed in urban environments. Urban habitats are typically harsher, with less and/or new resources to explore, also containing new unknown threats and a highly changing environment (Wheater, 1999). Hence, urban dwellers may need to have better cognitive abilities and flexibility in their behavior than forest or agricultural bees.

Some examples of behaviours needed to survive in urban environments are the use of human-made materials to nest (Allasino et al., 2019) or being able to collect pollen and nectar from exotic flowers from gardens (Lowenstein et al., 2014). We should not expect forested habitats to require such a plastic behaviour because the resources found in these habitats are homogeneous, stable, and bees are already pre-adapted to use them (Smith et al., 2019). In addition, other traits have already been described to be filtered out in cities, such as smaller body sizes (REF), early spring species or brood parasites (Harrison et al., 2017). We also found small species to be less frequent in urban habitats, maybe as a consequence of the patchy resource distribution, which need from larger body sizes, which confer larger foraging ranges, to be accessed [@kendall2019].

Despite brain weight is relatively easy to measure and standardize with the used protocol, recent studies challenge the use of brain size as an estimator for behavioral complexity and computational power (Chittka and Niven, 2009, and references therein). However, many studies have shown correlations between brain size and performance (Deaner et al., 2007; Gronenberg and Couvillon, 2010; Lefebvre and Sol, 2008), and brain size is correlated with learning abilities in bees (Collado et al 2021). Finer scale measures such as neuropils size or detailed mushroom bodies measurements (the suspected centers of cognitive processes in bees; Barth & Heisenberg, 1997; Withers et al., 1995; Fahrbachand & Robinson 1995) could further explain differences in behaviours, and in consequence, shed extra light on species habitat associations. In any case, raw measures of brain size have the advantage to allow for larger comparisons across species.

@@

@@

>>>>

---
title: "Bees need larger brains to thrive in urban environmnets"
output:
  pdf_document:
    fig_caption: yes

@@

header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float} 
- \floatplacement{figure}{H}

csl: BiologyLetters.csl

---

@@

@@

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role in ecosystems for fruit and seed production. However, not all pollinators are negatively affected by habitat conversion, but some find in anthropogenic landscapes appropriate resources to survive and reproduce. Why some animals can thrive in anthropogenic environments that most species cannot tolerate remains insufficiently understood, but it might in part reflect enhanced behavioural plasticity to exploit new resource opportunities. While this hypothesis has received ample support in studies of vertebrates, it is less obvious whether the small brain of insects can offer enough behavioural plasticity to cope with the many challenges of anthropogenic habitats. To address this issue, we measured brains for `r sum(d$individuals)` individuals from `r nlevels(factor(unique(d$Species)))` species of bees, the most important group of pollinators, and compared them with detailed information on habitat preferences. Our analyses revealed that bees that prefer forested or agricultural habitats had smaller brains relative to their body size, but those who prefer urban habitats tended to have larger relative brain size. In addition, urban bees also tend to be larger, and hence, show larger absolute brain sizes. These results suggest that a large brain can help maintaining bee urban populations.

Keywords: relative brain size, habitat preference, Apoidea, urbanization.

Here, we characterize habitat preference for `r nlevels(factor(unique(d$Species)))` bee species and relate this to their brain size, expecting to find bigger brain sizes for urban dwellers because a wider array of cognitive abilities are needed to inhabit these habitats. In vertebrates, relative brain size is a good proxy of problem-solving and learning (REF!!), which can help individuals to cope with the many challenges on novel environments. Despite there was no unequivocal evidence that insect brain size is related to cognitive performance, learning abilities across bee species are reflected in differences in their brain size [@collado2021].  Hence, our study explores the long standing idea that species with higher encephalisation (i.e., large brain mass relative to their body) are more likely to thrive in urbanised areas applies also to insects.

@@

Our dataset contains measurements of brain and body size for bee specimens captured by hand netting in different areas of the East Coast of United States and Europe (i.e., Spain and Netherlands). In total, the dataset includes `r sum(d$individuals)` female individuals from `r nlevels(factor(unique(d$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` different families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) females are involved in a larger number of tasks than males and are likely to experience greater selective pressures from the environment; and (ii), there are evident behavioral and morphological differences between sexes that are associated with the size and structure of the brain [@streinzer2013; @roselino2015]. In addition, because brain size can change with age and experience of the different individuals within a species [@durst1994; @withers2008], we limited our sampling to specimens foraging on flowers. Note that 93 of the species used in our study were previously published with their respective trait measurements in @sayol2020. Detail information of how brain and body size were measured can be found in supplementary material.

## (b) Occurrence data

@@

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. For North America, we selected states located on the East Coast of United States (Figure S1A) that covered an approximate area of `r usakm2` km$^{2}$.  Similarly, we selected countries located on the north and center of Europe (Figure S1B) that represented a total area of `r eukm2` km$^{2}$. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a distribution larger than the area selected; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large differences with the extracted land cover classifications, which is from 2006.

## (c) Land cover classification

@@

## (d) Construction of the phylogenetic tree

@@

## (e) Analysis

We estimated brain size relative to body size modeling the log-transformed association between brain weight and intertegular span. For this, we specified a simple Bayesian regression with the function *brm* from the package *brms* version 2.18 [@burkner2017] where brain weight was the response variable and intertegular span the predictor. To correct for the non-independent evolutionary histories among species, we included the phylogenetic covariance matrix as random factor (see supplementary text X for phylogenetic tree construction). Note that this matrix contained all species of our dataset (N = `r spp_number_nofilter`) because we expect better predictive power with a larger sample size than the one with the filtered dataset (N = `r spp_number`). Lastly, we extracted the differences between the observed and predicted values (i.e., residuals) that represent the relative brain size. High values indicate larger brains than expected for their body size and low values denote smaller brains than expected.

@@

All our analyses were undertaken in R version 4.0.5 [@r2021], and all data processing and graphics were done with the set of packages from the tidyverse version 1.3.0 [@tidyverse19].

@@

Bees with a preference for inhabiting urban habitats tended to have bigger brains related to their body sizes, and those with forest or agricultural  preferences had relatively smaller brains. This pattern is highly suggestive, because investing in brain tissue is expensive (REF), and hence it only payoff in unpredictable habitats where a flexible behavior is needed. When the environment is stable, small relative brains are cost-effective to exploit predictable resources, but in highly unpredictable environments, a large brain can confer the needed flexibility to exploit unpredictable resources. Indeed, specialist bees also tend to have larger brains [@sayol2020], as they need to locate and memorize patchy resources. This same complex learning and navigation skills may be also needed in urban environments. Urban habitats are typically harsher, with less and/or new resources to explore, also containing new unknown threats and a highly changing environment [@wheater2002]. Hence, urban dwellers may need to have better cognitive abilities and flexibility in their behavior than forest or agricultural bees.

Some examples of behaviours needed to survive in urban environments are the use of human-made materials to nest [@allasino2019] or being able to collect pollen and nectar from exotic flowers from gardens [@lowenstein2019]. We should not expect forested habitats to require such a plastic behaviour because the resources found in these habitats are homogeneous, stable, and bees are already pre-adapted to use them [@smith2019]. In addition, other traits have already been described to be filtered out in cities, such as smaller body sizes (REF), early spring species or brood parasites [@harrison2019]. We also found small species to be less frequent in urban habitats, maybe as a consequence of the patchy resource distribution, which need from larger body sizes, which confer larger foraging ranges, to be accessed [@kendall2019].

Despite brain weight is relatively easy to measure and standardize with the used protocol, recent studies challenge the use of brain size as an estimator for behavioral complexity and computational power [@chittka2009\; and references therein]. However, many studies have shown correlations between brain size and performance [@deaner2007; @gronenberg2010; @sol2008], and brain size is correlated with learning abilities in bees [@collado2021]. Finer scale measures such as neuropils size or detailed mushroom bodies measurements [the suspected centers of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] could further explain differences in behaviours, and in consequence, shed extra light on species habitat associations. In any case, raw measures of brain size have the advantage to allow for larger comparisons across species.

@@

# Data accessibility

All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.5s5kuf. In addition, bee trait data and code can be found in the online repositories of Zenodo XXXXXXXKXX and Github https://github.com/ibartomeus/declines_brain.git.

# Authors' contributions

JBL, MAC, FS, IB...

# Funding

# Acknowledgements

# Conflict of interest declaration

We declare we have no competing interests.

@@

## Brain and body size measurements

Brain size was measured as the weight of the fixed brain for each specimen. First, all collected individuals were kept in cold and sent to the laboratory where they were identified by expert taxonomists and anesthetized in cold. Then, heads were removed with a scalpel and placed in a fixative solution (4% paraformaldehyde in 0.1 M phosphate-buffered saline -PBS- pH = 7.4) to avoid degradation. Subsequently, brains were extracted from the head capsule and placed on a petri dish after removing the retina from the optic lobes and cleaning the entire brain of all tracheae and fat bodies. By doing this, we ensured that our brain measurements mainly accounted for neural tissue. We then placed the brains on a small piece of Parafilm® and dried the exceeding fixative solution with Kimwipes® tissues. Then, within 4 seconds of the liquid removal, we weighted the brains with the help of a microbalance Sartorius Cubis®. Finally, we calculated the average weight for the species with more than one specimen. Because weighing bee brains is a highly error-prone task, we excluded values that were under or above 1.5 times the interquartile range for each species.

We considered as a proxy of body size the average intertegular span (ITS) per species. ITS represents the distance between the base of the wings (tegulae) on the bee thorax and has been shown to accurately predict body size as it is highly correlated with dry body weight (Cane et al., 2006; Kendall et al., 2019). All ITS measurements were conducted with a stereomicroscope (magnification 16x or 80x) with a calibrated ocular micrometre (resolution down to 0.02 mm).

Cane, J. H., Minckley, R. L., Kervin, L. J., Roulston, T. A. H., & Williams, N. M. (2006). Complex responses within a desert bee guild (Hymenoptera: Apiformes) to urban habitat fragmentation. _Ecological applications_, _16_(2), 632-644.

Kendall, L. K., Rader, R., Gagic, V., Cariveau, D. P., Albrecht, M., Baldock, K. C., ... & Bartomeus, I. (2019). Pollinator size and its consequences: Robust estimates of body size in pollinating insects. _Ecology and Evolution_, _9_(4), 1702-1714.
